---
title: "Basic R: Part 1"
author: "Nick Salkowski"
date: "Wednesday, February 25, 2015"
output: ioslides_presentation
---
  
## The Introduction
  
```{r setup, echo = FALSE}
knitr::opts_knit$set(tidy = FALSE)
```

There are a lot of sources of information about R.  An excellent, and useful place to start is *An Introduction to R* 
  
(<http://cran.r-project.org/doc/manuals/R-intro.pdf>)

## Case Sensitivity

- R is case sensitive.

* So, `X` and `x` are not the same.
* Get used to it.

## Objects

- R is object oriented.
- So, you will work with different kinds of objects, not just data tables.
- The most common objects are different kinds of vectors, matrices, arrays, lists, data.frames, and functions.
- Note that there are no scalars, only vectors of length 1.

## Functions

Functions are usually called using a function name followed by parentheses.  For example, the `c()` function combines values into a vector:

```{r c}
c(1, 2, 3, 4)
```

The `mean()` function calculates the mean:

```{r mean}
mean(c(1, 2, 3, 4))
```

## More Functions

R is full of functions.  Some things that don't *look* like functions are really functions, like `+()`:

```{r functions2}
5 + 3
`+`(2, 9)
```

## Arithmetic

- `+` Addition
- `-` Subtraction
- `*` Multiplication
- `/` Division
- `^` Exponentiation
- `%/%` & `%%` Integer Division & Remainder

## Arithmetic (continued)

```{r}
13 + 2
13 - 2
13 * 2
13 / 2
```

## Arithmetic (continued)

```{r}
13 ^ 2
13 %/% 2
13 %% 2
```

## Relational Operators

- `<` Less Than
- `>` Greater Than
- `<=` Less Than or Equal To
- `>=` Greater Than or Equal To
- `==` Equal To
- `!=` Not Equal To

## Relational Operators (continued)

```{r}
13 < 2
13 > 2
13 == 2
13 != 2
```

## Membership

`%in%` is used to test for membership.

```{r}
"A" %in% c("A", "B", "C")
c("A", "D") %in% c("A", "B", "C")
16 %in% c(1, 4, 9)
```

## Logical Operators

- `!` Not
- `&` And
- `|` Or

```{r}
5 > 2 & 10 < 3
5 > 2 | !(10 > 3)
```

## any() and all()

- `any()` tests whether there are any `TRUE` values
- `all()` tests whether all the values are `TRUE`

```{r}
any(c(TRUE, TRUE, FALSE))
all(c(TRUE, TRUE, FALSE))

```

## NA and NULL

- `NA` means "missing"
- `NULL` means "nothing"
- Use `is.na()` to test whether a value is missing
- Use `is.null()` to test whether a value is nonexistant

## NA and NULL (continued)

```{r}
c(TRUE | NA, FALSE | NA)
c(all(c(TRUE, NA)), all(c(TRUE, NULL)))
is.na(c("A", NA))
is.null(c("A", NULL))
```

## Assignment Operators

To assign a name to an object, use an assignment operator:

```{r assignment}
X <- 5
X
X * 2 -> Y
Y
```

## More Assignment Operators

- `=` can be used instead of `<-`, but it isn't recommended!
- `<<-` and `->>` should be avoided until you understand *environments*.
- `assign()` can be useful within functions or loops.

```{r assign}
assign(x = "ABCDE", 
       value = c("A", "B", "C", "D", "E"))
ABCDE
```

## Assignment is a function, too

```{r assignfunc}
`<-`(XYZ, c("X", "Y", "Z"))
XYZ
```

## Floating Point Math

R is not immune from floating point math issues...

```{r}
P <- 0.1
Q <- 0.3
c(P == (Q / 3), (Q / P) == 3) # HUH?
c(P - (Q / 3), (Q / P) - 3)
```

## Vectors

R vectors come in a variety of types.  Here are some common types:

```{r vectors}
c("A", "B", "C") # character
c(0.1, 0.2, 0.3) # numeric
c(1L, 2L, 3L) # integer
```

## More Vectors

```{r morevectors}
c(TRUE, FALSE, TRUE) # logical
factor(c("A", "B", "C")) # factor
complex(
  real = c(1.1, 2.2, -3.3), 
  imaginary = c(0, -1, 4)) # complex
```

`raw` vectors also exist, but are rarely used.

## TRUE and FALSE

- When you start using R, there are two objects, `T` and `F`, which have values `TRUE` and `FALSE`, respectively.
- So, you may be tempted to just use `T` and `F` instead of `TRUE` and `FALSE` -- **DO NOT DO THIS!**
- The values of `T` and `F` may be replaced ...

```{r}
c(T, F)
T <- FALSE
F <- TRUE
```

## TRUE and FALSE (continued)

```{r}
T == FALSE
F == TRUE
```


## Coercion

Usually R knows what to do with two objects of the same type, but if you have two objects of different types, then R will usually try to *coerce* the objects to allow your function to work.  Consider the `c()` function.

```{r}
c(c("A", "B"), # character
  c(1.0, 0.2)) # numeric
c(c("A", "B"), # character
  c(1L, 2L))   # integer
```

## Coercion (continued)

```{r}
c(c("A", "B"),    # character
  c(TRUE, FALSE)) # logical
c(c("A", "B"),         # character
  factor(c("a", "b"))) # factor
```

## factors

factors are *strange* objects, and cause all sorts of trouble if you don't know what you're doing.

- factors *look* like character vectors, but deep down they're really *integer* vectors, with an *attribute* called `levels` that maps the integers to character strings.
- So, factor vectors behave *really differently* than character vectors.
- Also, R has a habit of converting character vectors into factor vectors by default, so factors can be easily generated if you don't pay attention.

## Coercion (continued)

```{r}
c(c(0.1, 0.2),    # numeric
  c(1L, 2L))      # integer
c(c(0.1, 0.2),    # numeric
  c(TRUE, FALSE)) # logical
c(c(1L, 2L),      # integer
  c(TRUE, FALSE)) # logical
```

## Coercion (continued)

```{r}
"1" == 1.0
"1" == 1L
1.0 == 1L
```

## Vector Arithmetic with Scalars

Really, Vector Arithmetic with Vectors with length 1!
  
```{r vectormath}
X <- c(1, 2, 3, 4, 5)
X + 10
X * 2
1 / X
```

## Vector Arithmetic with Vectors

```{r vectormath2}
Y <- c(1, -2, 3, -4, 5)
X + Y
X * Y
X %*% Y
```

## Outer Products

```{r outer}
outer(X, Y)
```

## More Outer Products
```{r outer2}
outer(X, Y, "+")
```

## Getting Help

There are lots of ways to access help:
  
- `?sample`
- `help(sample)`

You can also search the help for a particular string:
  
- `help.search("sample")`
- `??sample`