---
title: "The Basics"
author: "Nick Salkowski"
date: "Thursday, February 26, 2015"
output: ioslides_presentation
---

## The Introduction

```{r, echo = FALSE}
knitr::opts_knit$set(tidy = FALSE)
```

There are a lot of sources of information about R.  A useful place to start is **An Introduction to R**.

<http://cran.r-project.org/doc/manuals/R-intro.pdf>

## Case Sensitivity

- R is case sensitive.

    * So, `X` and `x` are not the same.
    * Get used to it.
    
## Objects

- R is object oriented.
- So, you will work with different kinds of objects, not just data tables.
- The most common objects are different kinds of vectors, matrices, arrays, lists, data.frames, and functions.
    - Note that there are no scalars, only vectors of length 1.

## Functions

- Functions are usually called using a function name followed by parentheses.
- Many (but not all) functions have arguments.  The arguments modify what the function does.
- Functions often **return** some result.
- e.g., `typeof()` is a function with one argument called `x`.
    - `x` is an object
    - `typeof` returns what **type** of object `x`
- Nearly everthing that **does something** in R is a function, even if it doesn't always **look like** a function.

## Data Types

```{r}
typeof(x = "A")
typeof(x = 1.2)
typeof(x = 1L)
typeof(x = TRUE)
```

## Function Arguments

- Some functions have more than one argument.
    - You can provide the arguments in any order you want, if you use the argument names.
    - You can omit the names, if you provide the arguments **in the order the function expects them!**
- e.g., `rep.int()` is a function with two arguments, `x` and `times`.
    - `x` is a vector.
    - `times` is a vector indicating the number of times to replicate each element of `x`.
    
## Replication

```{r}
rep.int(x = 5.1, times = 3)
rep.int(times = 3, x = 5.1)
rep.int(5.1, 3)
```

## Getting Help

The documentation for each function provides a desription of the function, including information about the arguments and what the function returns.

There are lots of ways to access documentation ("help files"):

```{r, eval = FALSE}
?rep.int
```
```{r, eval = FALSE}
help(rep.int)
```

You can also search the help for a particular string:

```{r, eval = FALSE}
??replicate
```
```{r, eval = FALSE}
help.search("replicate")
```

## Is `?` a function?

- `?` accesses documentation, so it certainly **does something**.
- `?` **is** a function, but a function with an unusual syntax.
- You **can** call `?` using the "normal" function syntax, too.  It takes more keystrokes, so the "normal" syntax is rarely used.

```{r eval = FALSE}
`?`(rep.int)
```

- Note the tick marks around the function name. Unusual function names need them.

## Constructing Vectors

- You can combine multiple values together into a vector using the `c()` function.

```{r}
c("A", "B", "C", "D")
c(1.2, 3.4)
c(TRUE, FALSE, TRUE)
```

## Vector Types

- Most vectors can only contain a single type of data.  
    - Character vectors can only contain character strings
    - Numeric vectors can only contain double-precision numbers
    - Integer vectors can only contain integers
    - Logical vectors can only contain TRUE and FALSE

## NA means "Missing"

- `NA` is a special value that means "missing"
- `is.na` is a function that tests whether values are missing

```{r}
c(1, 2, NA, 4)
is.na(c(1, 2, NA, 4))
```

## NULL means "Nothing"
- `NULL` is a special value that means "nothing"
- `is.null()` to test whether a value is nonexistant

```{r}
c("A", "B", NULL, "D")
is.null(c("A", "B", NULL, "D"))
```

## Assignment Operators

To assign a name to an object, use an assignment operator:

```{r}
X <- 1.2
3.4 -> Y
c(X, Y)
```

- Assignment is a function, too.

```{r}
`<-`(XYZ, c("X", "Y", "Z"))
XYZ
```

## More Assignment Operators

- `=` can be used instead of `<-`, but using `=` can be confusing, since `=` is also used for function arguments.
- `<<-` and `->>` should be avoided until you understand *environments*.
- `assign()` can be useful within functions or loops.
```{r assign}
assign(x = "ABCDE", 
       value = c("A", "B", "C", "D", "E"))
ABCDE
```

## Constructing Matrices

- You can turn a vector into a matrix using the `matrix` function.

```{r}
matrix(
  data = c(1, 2, 3, 4, 5, 6),
  nrow = 2)
```

- Note that the matrix was "filled" by columns. The `byrow` argument to `matrix` can be used to control this behavior.

## Lists

- Lists are special vectors whose elements can be different.
    - Elements can be different data types.
    - Elements can be complex objects, not just single values.
- Lists are constructed using the `list()` function

```{r}
list("A", 1.2, TRUE)
```

## Complex List Elements

```{r}
list(c("A", "B"), c(1L, 2L, 3L), matrix(c(1.2, 3.4), nrow = 2))
```

## Naming List Elements

```{r}
list(AB = c("A", "B"), X = c(1L, 2L, 3L), 
     mat = matrix(c(1.2, 3.4), nrow = 2))
```

## data.frames

- A **data.frame** is a special kind of list, where all the elements of the list are vectors with the same length.  
- Because the vectors all have the same length, a data.frame has rows and columns, like a matrix, but each column can have different kinds of data, like a list.
- data.frames are constructed using the `data.frame()` function

```{r}
data.frame(ABC = c("A", "B", "C"), X = c(1L, 2L, 3L), 
           TF = c(TRUE, FALSE, NA))
```

## Recycling Vectors

- When you supply a vector that is too short, R often "recycles" its values to get a vector of the necessary length.

```{r}
matrix(data = c(1, 2, 3, 4, 5), nrow = 4, ncol = 5)
```

## Extracting Elements from Objects

- R has several different ways to extract elements from objects.  
- Not every technique works with every object!
- So, familiarity with a variety of methods is important.
    - Bracket Notation
    - Double Bracket Notation
    - Dollar Sign Notation
    
## Bracket Notation (1)

- Bracket notation can use integer indices, element names, or logical values.
- Positive integer indices indicate which elements to retain.

```{r}
ABCDE <- c("A", "B", "C", "D", "E")
ABCDE[2]
ABCDE[c(1, 3, 4)]
```

## Bracket Notation (2)

- Positive integer indices can be repeated.
```{r}
ABCDE[c(1, 1, 3, 4, 3)]
```

- Negative indices indicate which elements to drop.
```{r}
ABCDE[c(-2, -4)]
```

## Bracket Notation (3)

- Named elements can be extracted using character strings.

```{r}
Friends <- c(Ann = 1L, Bob = 2L, Cathy = 3L, Doug = 4L)
Friends["Ann"]
Friends[c("Bob", "Doug")]
```

## Bracket Notation (4)

- A vector of logical values can be used for extraction.
```{r}
ABCDE <- c("A", "B", "C", "D", "E")
ABCDE[c(TRUE, TRUE, FALSE, FALSE, TRUE)]
```

- Recycling!

```{r}
ABCDE <- c("A", "B", "C", "D", "E")
ABCDE[c(TRUE, FALSE, TRUE)]
```

## Bracket Notation (5)

- Matrices have two dimensions, so the bracket notation controls both rows and columns.
- `LETTERS` and `letters` are character vectors of the upper- and lowercase letters of the alphabet.

```{r}
LETTERS12 <- matrix(LETTERS[1:12], nrow = 3)
LETTERS12
```

- Note, `1:12` is R shorthand for an vector of integers from 1 to 12.

## Bracket Notation (6)

- Extract a row from a matrix:

```{r}
LETTERS12[3, ]
```

- Drop a row from a matrix:
```{r}
LETTERS12[-3, ]
```

## Bracket Notation (7)

- Extract some columns from a matrix:

```{r}
LETTERS12[, c(2, 4)]
```

- Extract from both rows and columns:

```{r}
LETTERS12[3, c(2, 4)]
```

## Bracket Notation (8)

- If the matrix has row or column names, character strings can be used, too.
```{r}
INT6 <- matrix(1:6, nrow = 2, 
                   dimnames = list(c("AM", "PM"), 
                                  c("Mon", "Tue", "Wed")))
INT6
INT6["AM", c("Mon", "Wed")]
```


